---
title: "Application-based simulation setup"
format: html
---

```{r}
.libPaths("~/apps/R_4.1.0/")
library(NMF)
library(bayesNMF)
library(tidyverse)
library(knitr)
library(glue)

knitr::opts_chunk$set(dev = "ragg_png", echo = TRUE, fig.show = "hold")

cosmic <- bayesNMF::get_cosmic()
application_dir <- "."
figures_dir <- "application_based_simulations_figures"
```

## Load Breast AdenoCarcinoma Data

```{r}
M <- read.csv(file.path(application_dir, "M_BA.csv"))
Tr <- read.csv(file.path(application_dir, "T_BA.csv"))
Tr_sub45 <- read.csv(file.path(application_dir, "T_BA_sub45.csv"))

M_sub45 <- M[,paste0("Breast.AdenoCA..", Tr_sub45$tumor_wgs_icgc_specimen_id)]

head(Tr)
```

```{r}
untreated_IDs <- Tr %>% 
  filter(BA_T == FALSE) %>%
  mutate(
    # to match colnames of M
    tumor_wgs_icgc_specimen_id = paste0("Breast.AdenoCA..", tumor_wgs_icgc_specimen_id) 
  ) %>%
  pull(tumor_wgs_icgc_specimen_id)

M_untreated <- M[, untreated_IDs]
dim(M_untreated)
```

## Choosing Latent Rank

We perform the same rank choosing strategies on all data combined and on untreated data only. Both yield the same conclusion with a **selected rank N = 5**. This rank balances optimizing traditional NMF metrics as well as BIC assuming a Poisson likelihood.

```{r}
nmf_rescale_align <- function(M, N, nrun = 10) {
  nmf_out <- NMF::nmf(M, rank = N, nrun = nrun)
  Phat <- nmf_out@fit@W
  Ehat <- nmf_out@fit@H
  
  Ehat <- sweep(Ehat, 1, colSums(Phat), '*')
  Phat <- sweep(Phat, 2, colSums(Phat), '/')
  stopifnot(all(abs(colSums(Phat) - 1) < 0.001))
  
  sim_mat <- bayesNMF::pairwise_sim(Phat, cosmic)
  assigned <- bayesNMF::assign_signatures(sim_mat)
  
  colnames(Phat) <- colnames(assigned)
  rownames(Ehat) <- colnames(assigned)

  return(list(
    nmf = nmf_out,
    Phat = Phat,
    Ehat = Ehat,
    sim_mat = assigned
  ))
}

plot_Ehat <- function(Ehat) {
  t(Ehat) %>% 
    data.frame() %>%
    pivot_longer(1:N, names_to = 'sig', values_to = 'estimated_C') %>%
    ggplot(aes(x = estimated_C)) +
    facet_grid(rows = vars(sig), switch = 'y') +
    geom_histogram() +
    theme(
      text = element_text(size = 30),
      strip.text.y.left = element_text(angle = 0)
    )
}
```

### Using all data

Run NMF for ranks 1-15 with 10 runs per rank and plot metrics

```{r}
set.seed(321)
rank_sel_nmf <- NMF::nmf(M, rank = 1:15, nrun = 10)

png(file.path(figures_dir, "rank_sel_nmf_BA.png"), width = 800, height = 600)
plot(rank_sel_nmf)
dev.off()

BICs <- sapply(1:15, function(rank) {
  Phat = rank_sel_nmf$fit[[as.character(rank)]]@fit@W
  Ehat = rank_sel_nmf$fit[[as.character(rank)]]@fit@H
  Mhat = Phat %*% Ehat
  loglik = sum(sapply(1:ncol(M), function(g) {
    sum(dpois(M[,g], Mhat[,g], log = TRUE))
  }))

  k = rank*(nrow(M) + ncol(M)) # number of parameters 
  n = ncol(M)                  # number of samples
  BIC = k*log(n) - 2*loglik
  return(BIC)
})
png(file.path(figures_dir, "BICs_BA.png"), width = 800, height = 600)
plot(BICs, xlab = "rank", ylab = "BIC")
lines(BICs)
abline(v = which(BICs == min(BICs)))
dev.off()

which(BICs == min(BICs))

which(BICs < min(BICs)*1.05)
which(BICs < min(BICs)*1.01)
```

Optimizing BIC gives N = 7, but many ranks are very close (\<5% increase away) to the minimum. Optimizing cophenetic and basis sparsity gives N = 4 or 5. Dispersion and silhouette aren't great for N = 4 or 5 but there are sharp drops right after it. This is also a second elbow of evar and rss.

```{r}
MaxCos <- sapply(1:15, function(rank) {
  if (rank == 1) {
    return(0)
  }
  Phat = rank_sel_nmf$fit[[as.character(rank)]]@fit@W
  sim = pairwise_sim(Phat, Phat)
  max_cos = max(sim[lower.tri(sim)])
  return(max_cos)
})
png(file.path(figures_dir, "MaxCos_BA.png"), width = 800, height = 600)
plot(MaxCos, xlab = "rank", ylab = "Max Cosine Sim")
lines(MaxCos)
abline(h = 0.7)
dev.off()
```

This plot shows the maximum cosine similarity between two estimated signatures. We place a horizontal line at cosine similarity = 0.7, values above which indicate likely duplicate signatures. This seems to place a cap at a rank of N = 5.

For N = 4, 5, 7, run NMF again, rescale P to sum to 1, align to COSMIC, and plot exposures per signature

```{r}
N = 4
set.seed(321)
nmf_out <- nmf_rescale_align(M, N)
nmf_out$sim_mat

get_heatmap(nmf_out$Phat, cosmic)
ggsave(file.path(figures_dir, "heatmap_N4_BA.png"))

plot_Ehat(nmf_out$Ehat)
ggsave(file.path(figures_dir, "Ehat_N4_BA.png"))

BICs[4]
(BICs[4] - BICs[7])/BICs[7]
```

N = 4 has one signature not well aligned to COSMIC

```{r}
N = 5
set.seed(321)
nmf_out <- nmf_rescale_align(M, N)
nmf_out$sim_mat

get_heatmap(nmf_out$Phat, cosmic)
ggsave(file.path(figures_dir, "heatmap_N5_BA.png"))

plot_Ehat(nmf_out$Ehat)
ggsave(file.path(figures_dir, "Ehat_N5_BA.png"))

BICs[5]
(BICs[5] - BICs[7])/BICs[7]
```

N = 5 has better alignment to COSMIC, and BIC is within 1.5% of minimum

```{r}
N = 6
set.seed(321)
nmf_out <- nmf_rescale_align(M, N)
nmf_out$sim_mat

get_heatmap(nmf_out$Phat, cosmic)
ggsave(file.path(figures_dir, "heatmap_N6_BA.png"))

plot_Ehat(nmf_out$Ehat)
ggsave(file.path(figures_dir, "Ehat_N6_BA.png"))

BICs[6]
(BICs[6] - BICs[7])/BICs[7]
```

```{r}
N = 7
set.seed(321)
nmf_out <- nmf_rescale_align(M, N)
nmf_out$sim_mat

get_heatmap(nmf_out$Phat, cosmic)
ggsave(file.path(figures_dir, "heatmap_N7_BA.png"))

plot_Ehat(nmf_out$Ehat)
ggsave(file.path(figures_dir, "Ehat_N7_BA.png"))

BICs[7]
```

N = 7 seems to have a duplicate signature (SBS3 and 40)

### Using untreated samples only

Run NMF for ranks 1-15 with 10 runs per rank and plot metrics

```{r}
set.seed(321)
rank_sel_nmf_untreated <- NMF::nmf(M_untreated, rank = 1:15, nrun = 10)

png(file.path(figures_dir, "rank_sel_nmf_untreated_BA.png"), width = 800, height = 600)
plot(rank_sel_nmf_untreated)
dev.off()

BICs_untreated <- sapply(1:15, function(rank) {
  Phat = rank_sel_nmf_untreated$fit[[as.character(rank)]]@fit@W
  Ehat = rank_sel_nmf_untreated$fit[[as.character(rank)]]@fit@H
  Mhat = Phat %*% Ehat
  loglik = sum(sapply(1:ncol(M_untreated), function(g) {
    sum(dpois(M_untreated[,g], Mhat[,g], log = TRUE))
  }))

  k = rank*(nrow(M_untreated) + ncol(M_untreated)) # number of parameters 
  n = ncol(M_untreated)                            # number of samples
  BIC = k*log(n) - 2*loglik
  return(BIC)
})
png(file.path(figures_dir, "BICs_untreated_BA.png"), width = 800, height = 600)
plot(BICs_untreated, xlab = "rank", ylab = "BIC")
abline(v = which(BICs_untreated == min(BICs_untreated)))
lines(BICs_untreated)
dev.off()

which(BICs_untreated == min(BICs_untreated))

which(BICs_untreated < min(BICs_untreated)*1.05)
which(BICs_untreated < min(BICs_untreated)*1.01)
```

```{r}
MaxCos_untreated <- sapply(1:15, function(rank) {
  if (rank == 1) {
    return(0)
  }
  Phat = rank_sel_nmf_untreated$fit[[as.character(rank)]]@fit@W
  sim = pairwise_sim(Phat, Phat)
  max_cos = max(sim[lower.tri(sim)])
  return(max_cos)
})
png(file.path(figures_dir, "MaxCos_untreated_BA.png"), width = 800, height = 600)
plot(MaxCos_untreated, xlab = "rank", ylab = "Max Cosine Sim")
lines(MaxCos_untreated)
abline(h = 0.7)
dev.off()
```

Same as before!

For N = 4, 5, 7, run NMF again, rescale P to sum to 1, align to COSMIC, and plot exposures per signature

```{r}
N = 4
set.seed(321)
nmf_out <- nmf_rescale_align(M_untreated, N)
nmf_out$sim_mat

get_heatmap(nmf_out$Phat, cosmic)
ggsave(file.path(figures_dir, "heatmap_N4_untreated_BA.png"))

plot_Ehat(nmf_out$Ehat)
ggsave(file.path(figures_dir, "Ehat_N4_untreated_BA.png"))

BICs_untreated[4]
(BICs_untreated[4] - BICs_untreated[7])/BICs_untreated[7]
```

N = 4 has one signature not well aligned to COSMIC

```{r}
N = 5
set.seed(321)
nmf_out <- nmf_rescale_align(M_untreated, N)
nmf_out$sim_mat

get_heatmap(nmf_out$Phat, cosmic)
ggsave(file.path(figures_dir, "heatmap_N5_untreated_BA.png"))

plot_Ehat(nmf_out$Ehat)
ggsave(file.path(figures_dir, "Ehat_N5_untreated_BA.png"))

BICs_untreated[5]
(BICs_untreated[5] - BICs_untreated[7])/BICs_untreated[7]
```

N = 5 has better alignment to COSMIC, and BIC is within 1.5% of minimum

```{r}
N = 6
set.seed(321)
nmf_out <- nmf_rescale_align(M_untreated, N)
nmf_out$sim_mat


get_heatmap(nmf_out$Phat, cosmic)
ggsave(file.path(figures_dir, "heatmap_N6_untreated_BA.png"))

plot_Ehat(nmf_out$Ehat)
ggsave(file.path(figures_dir, "Ehat_N6_untreated_BA.png"))

BICs_untreated[6]
```

```{r}
N = 7
set.seed(321)
nmf_out <- nmf_rescale_align(M_untreated, N)
nmf_out$sim_mat


get_heatmap(nmf_out$Phat, cosmic)
ggsave(file.path(figures_dir, "heatmap_N7_untreated_BA.png"))

plot_Ehat(nmf_out$Ehat)
ggsave(file.path(figures_dir, "Ehat_N7_untreated_BA.png"))

BICs_untreated[7]
```

N = 7 seems to have a duplicate signature (SBS3 and 37)

## Fixing Signatures

We fix signatures as the reference COSMIC signatures that best align to the NMF results with chosen rank N = 5.

```{r}
N = 5
set.seed(321)
nmf_out <- nmf_rescale_align(M_untreated, N)

fixed_P = cosmic[,colnames(nmf_out$P)]
colnames(fixed_P)
```

## Sampling Distribution of Untreated Contributions

```{r}
fixed_C0_dist <- do.call(cbind, lapply(1:ncol(M_untreated), function(g) {
  nnls::nnls(A = fixed_P, b = M_untreated[,g])$x
}))
rownames(fixed_C0_dist) = colnames(fixed_P)
dim(fixed_C0_dist)

plot_Ehat(fixed_C0_dist)
ggsave(file.path(figures_dir, "fixed_C0_dist_BA.png"), height = 13, width = 8)
```

```{r}
sd_above = 4

todrop = rep(0, ncol(fixed_C0_dist))
for (n in 1:N) {
  print(colnames(fixed_P)[n])
  upper = mean(fixed_C0_dist[n,]) + sd_above*sd(fixed_C0_dist[n,])
  todrop = todrop + (fixed_C0_dist[n,] > upper)
}
mean(todrop > 0)
sum(todrop > 0)

plot_Ehat(fixed_C0_dist[,todrop == 0])
ggsave(file.path(figures_dir, "fixed_C0_dist_filtered_BA.png"), height = 13, width = 8)
```

## Save simulation parameters to .rds file

```{r}
simulation_parameters <- list(
  P = fixed_P, 
  C0_dist = fixed_C0_dist, 
  ATE_C = ATE_C, 
  ATE_C_sd = ATE_C_sd
)
saveRDS(simulation_parameters, file = file.path(application_dir, "simulation_parameters_BA.rds"))
```
